---
title: "Analysis of COVID-era Apple Mobility Data"
author: "Adam Zimmerman"
date: "9/28/2020"
output: html_document
bibliography: references.bib
params:
  state: "South Carolina"
  data: "data/raw_data/applemobilitytrends-2020-10-10.csv"
  seqdata: "output/covid_sequence_tallies.txt"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Setup

```{r set-parameters}
# set the state to work with
state_to_analyze <- params$state

# sanitizing the state name for output files
state_no_spaces <- gsub(state_to_analyze,
                        pattern = " ",
                        replacement = "_")

# set the name of the global data file to use
global_file_to_process <- params$data

# extract core name for output file naming
core_name <- tools::file_path_sans_ext(basename(global_file_to_process))
```

```{r load-packages-function}
# load the packages needed
library("dplyr")
library("readr")
library("ggplot2")
library("tidyr")
library("lubridate")
library("devtools")

# load the functions needed
source("code/functions/subset_mobility_data_to_state.R")
source("code/functions/count_counties_cities.R")
source("code/functions/flip_dataset.R")
```

## Background and Objectives

This markdown file uses `ggplot2` [@ggplot2] and `dplyr`[@dplyr] packages to analyze the Apple Maps mobility data from the time before the COVID-19 pandemic through the initial shutdowns and continuing changes over the course of the year.

The goal is to create a subset to a particular state, tally up the number of cities and counties in that state, and then to make several plots to visualize that data.

The state analyzed in this document is `r state_to_analyze`.

## Data Processing
```{r subset-state}
# subset out to the chosen state using the variables set above
state_subset <- subset_mobility_data_to_state(
  input_filename = global_file_to_process,
  subregion = state_to_analyze)
```

```{r make-subset-long}
# convert the subsetted state data from wide to long format
long_subset <- flip_dataset(
  file_name_in = paste0("output/subsetted_states_wide/",
                        state_no_spaces,
                        "_",
                        core_name,
                        ".csv"))
```

```{r tallied-subset}
# tally up the numbers of cities and counties in the state that have mobility
# data available
tallied_subset <- count_cities_counties(
  file_name_in = paste0("output/subsetted_states_wide/",
                      state_no_spaces,
                      "_",
                      core_name,
                      ".csv"))
```

## Figures

```{r plot-tallied}
# create a grouped barplot of the tallied data to include in the knitted
# document, and save that figure to the output/figures directory as a png
tallied_plot <- ggplot(data = tallied_subset,
                       aes(x = geo_type,
                       y = n,
                       fill = transportation_type)) +
  geom_col(position = position_dodge()) +
  labs(title = paste("Number of cities and/or counties in",
                     state_to_analyze,
                     "with mobility data"),
       x = "Type of area",
       y = "Number of areas")

ggsave(plot = tallied_plot,
       filename = paste0("output/figures/",
                         state_no_spaces,
                         "_",
                         core_name,
                         "_tallied_plot.png"))

tallied_plot
```

```{r plot-time-series}
# produce a line plot of relative driving mobility across the state with data
# on the x axis and mean relative mobility on the y axis
timeseries_plot <- long_subset %>%
  filter(transportation_type == "driving") %>%
  group_by(date) %>%
  summarize(mean_mobility = mean(rel_mobility)) %>%
  ggplot(aes(x = lubridate::ymd(date),
             y = mean_mobility)) +
  geom_line() +
  labs(title = paste("Statewide mean relative mobility driving levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

ggsave(plot = timeseries_plot,
       filename = paste0("output/figures/",
                         state_no_spaces,
                         "_",
                         core_name,
                         "_timeseries_plot.png"))

timeseries_plot
```

```{r plot-sequence-data, fig.height=12}
# read in the count of the sequences by country (2 columns)
seqs_by_country <- readr::read_table(params$seqdata,
                                     col_names = c("count",
                                                   "country"))
# check for two columns
stopifnot(ncol(seqs_by_country) == 2)

# make a sideways barplot, log scale x axis, with countries on the y
# and the counts of sequences in each on the x, sorted by count
seq_count_plot <- ggplot(data = seqs_by_country,
                         aes(x = count,
                             y = reorder(country, count))) +
  geom_col() +
  scale_x_log10() +
  labs(title = "Count of SARS-CoV-2 sequences per country",
       x = "Count of Sequences",
       y = "Country Name")

ggsave(plot = seq_count_plot,
       filename = "output/figures/seq_count_plot.png")

seq_count_plot

```

```{r plot-special}
# produce pre- and post- lockdown graphs of 3 lines each, one line for each
# transit type, with day of the week on the x axis and mean relative mobility
# on the y axis. The first graph should be from January and the second graph
# should be aggregate from March to April
lockdown_plot <- long_subset %>%
  group_by(date) %>%
  summarize(mean_mobility = mean(rel_mobility)) %>%
  ggplot(aes(x = lubridate::wday(date, label = TRUE),
             y = mean_mobility)) +
  geom_boxplot() +
  labs(title = paste("Weekly statewide mean relative mobility levels in",
                     state_to_analyze,
                     "before and during COVID"),
       x = paste("Day of the Week"),
       y = "Mean relative mobility") +
  facet_wrap(. ~ factor(ifelse(lubridate::month(date) < 3,
                        "Pre-COVID", # facet label for TRUE
                        "During COVID"), # facet label for FALSE
                        levels = c("Pre-COVID", # switch order so pre goes first
                                   "During COVID")))

ggsave(plot = lockdown_plot,
       filename = paste0("output/figures/",
                         state_no_spaces,
                         "_",
                         core_name,
                         "_lockdown_plot.png"))

lockdown_plot
```

## Session Info

```{r session-info}
devtools::session_info()
```

## Sources Cited
